---
title: "Class05: Data Viz with GGPLOT"
author: "Montserrat (PID:A16536527)"
format: pdf
---

Today we playing with plotting and graphics in E 

There are lots of ways to make cool figures in T 
There is "base" R graphics (`plot()`, `hist()`, `boxplot()` etc.)

there is also add-on packages like **ggplot**

```{r}
head(cars)
```
 lets plot with this "base" R: 
 
 option command i = insert code
 
```{r}
plot(cars)
```

```{r}
head(mtcars)
```
lets plot `mpg` vs `disp`
pch number can be changed to adjust figure for points 

```{r}
plot(mtcars$mpg,mtcars$disp,pch=12)
```
```{r}
hist(mtcars$mpg)
```

##GGPLOT

the main function in the ggplot2 package is **ggplot()*. First i need to install the ggplot2 package. i can install any package with the function `install.packages()` ( do this in the console)

>**N.B.** i never want to run `install.packages()` in my quarto source document

you need to load the package before attempting to use 

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)+
  aes(x=speed,y=dist)+
  geom_point()
```
every ggplot needas at least 3 things: data, 
= **data** (given with `ggplot(cars)`)
-the **aesthetic** mapping (given with `aes()`)
-the **geom** (given by `geom_point()`)

>for simple canned graphs "base" R is near always faster

```{r}
ggplot(cars)+
  aes(speed)+
  geom_histogram()
```


```{r}
ggplot(cars)+
  aes(x=speed,y=dist)+
  geom_point()+
  geom_line()
```

### add more layers

lets add a line and title, subtitle and caption as well as custom axis labels 

```{r}
ggplot(cars)+
  aes(x=speed,y=dist)+
  geom_point() + 
  geom_smooth()
```

straight line, no shading 

```{r}
ggplot(cars)+
  aes(x=speed,y=dist)+
  geom_point()+
  geom_smooth(method="lm",se=FALSE)+
  labs(title="Silly Plot", x="Speed(MPH)", y="Distance(ft)", caption="when is tea time anyway???") +
  theme_bw()
        
```
## Plot some expression data 

Read data file from online URL 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

>Q1. how many genes are in this wee dataset?

there are `r nrow(genes)` in this dataset 

>Q2. how may "up" regulated genes are there?

```{r}
sum(genes$State =="up")
```

there are `r sum(genes$State =="up")` UP genes

```{r}
table(genes$State)
```

```{r}
ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()
  
  
```


```{r}
ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=Condition2) +
    geom_point()
```

```{r}
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()+
  scale_color_manual(values=c("blue","gray","red"))
```

```{r}
p
```

```{r}
p+labs(title="look at me")
```

Silly example of adding labels 
```{r}
ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State,label=Gene) +
    geom_point()+
  scale_color_manual(values=c("blue","gray","red"))+ 
  geom_text()
```

```{r}
library(ggrepel)

ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State,label=Gene) +
    geom_point()+
  scale_color_manual(values=c("blue","gray","red"))+ 
  geom_text_repel(max.overlaps = 60)+ 
  theme_bw()
```

## Going Further 

playing with some different layers and the gapminder dataset

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
tail(gapminder)
```

A first plot 

```{r}
ggplot(gapminder)+
  aes(y=lifeExp,x=gdpPercap, col=continent)+
  geom_point()+
  facet_wrap(~continent)
```
alpha leads do transperency 
```{r}
ggplot(gapminder)+
  aes(y=lifeExp,x=gdpPercap, col=continent)+
  geom_point(alpha=0.2)+
  facet_wrap(~continent)+
  theme_bw()
```


```{r}
ggplot(gapminder)+
  aes(y=lifeExp,x=gdpPercap, col=continent, size=pop)+
  geom_point(alpha=0.2)+
  facet_wrap(~continent)+
  theme_bw()
```

###Extensions: Animation 

```{r}
library(gapminder)
library(gganimate)
```


```{r}
# Setup nice regular ggplot of the gapminder data
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  # Facet by continent
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  shadow_wake(wake_length = 0.1, alpha = FALSE)
```

###Combining Plots 

```{r}
library(dplyr)
library(patchwork)
```

```{r}
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))

(p1 | p2 | p3) /
      p4
```

