---
title: "Class 08: Breast Cancer Analysis Project"
author: "Montserrat (A16536527)"
format: pdf
toc: true
---

## Background 

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class. We will extend what you’ve learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.


## Data import 

data available on class website 
```{r}
#read.csv("WisconsinCancer.csv")
```

```{r}
# Save your input data file into your Project directory
fna.data <- read.csv("WisconsinCancer.csv")

# Complete the following code to input the data and store as wisc.df
wisc.df <- cbind(fna.data, row.names=1)

wisc.df
```
Make sure we do not include sample id or diagnosis columns in data that we analyze below 


```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
wisc.data <- wisc.df[,-1]
dim(wisc.data)

```

## Exploratory data analysis 

>Q1. How many observations are in this dataset?

There are `r nrow(wisc.data)`observations/samples/patients in data set 


>Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```
>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean",colnames(wisc.data)))
```

## Principal Component Analysis

The main function in base R for PCA is called `prcomp`. A optional argument `scale` should nearly always be switched to `scale=True` for this function 

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

always want to scale data so all of our columns are taken into consideration
```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp( wisc.data,scale=T )
summary(wisc.pr)


```

Lets make our main result figure - the "PC plot" or "score plot", "orientation plot"

```{r}
library(ggplot2)
ggplot(wisc.pr$x) + aes(PC1, PC2,col=diagnosis)+
         geom_point()
```

>Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

0.4427

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7

>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```
Difficult because everything is so clustered so close together 
```{r}
# Scatter plot observations by components 1 and 2
plot( wisc.pr$x , col = diagnosis , 
     xlab = "PC1", ylab = "PC2")
```
>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

PC1 and PC2 is more representative of the variance in the data compared to PC1 and PC3

```{r}
# Repeat for components 1 and 3
plot(wisc.pr$x[, -2 ], col = diagnosis, 
     xlab = "PC1", ylab = "PC3")
```
Make scatter plot colored by diagnosis 

```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```
## Variance explained 

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

## Communicating Results 

>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC.


```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled,method="euclidean")
wisc.hclust <- hclust(data.dist, method="complete")
```

>Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h= 19, col="red", lty=2)
```
## Selecting Number of Clusters 

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust,k=4)
table(wisc.hclust.clusters, diagnosis)
```

>Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

"ward.D2" method is my favorite it allows you to see the detect the number of clusters a bit better for this data 


## Comining PCA and clustering

```{r}
d <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(d,method="ward.D2")
plot(wisc.pr.hclust)
abline (h=70,col="red")
```
```{r}
grps<-cutree(wisc.pr.hclust,h=70)
table(grps)
```
Make a wee"cross-table"

```{r}
table(grps,diagnosis)
```
group 1 represents 24 benign and 179 malignant, group 2 represents 333 benign and 33 malignant suggesting that group 1  mainly is representative of malignant and group 2 of benign 

group 1: true positive (TP): 179, false positives (FP):24

Sensitivity: TP/(TP+FP)

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

```{r}
g <- as.factor(grps)
levels(g)
g <- relevel(g,2)
levels(g)
plot(wisc.pr$x[,1:2], col=g)
```

```{r}

wisc.pr.hclust.clusters <- cutree(wisc.hclust,k=4)
table(wisc.pr.hclust.clusters,diagnosis)
```

>Q13. How well does the newly created model with four clusters separate out the two diagnoses?

separates the clusters well 


>Q14. How well do the hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters,diagnosis)
```

Its does better 




