---
title: "Class 10: Halloween Mini-Project"
author: "Montserrat (PID:A16536627)"
format: pdf
toc: TRUE
---

## Importing Candy 

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
```{r}
flextable::flextable(head(candy))
```

>Q1. How many different candy types are in this dataset?

```{r}
library(dplyr)
nrow(candy)
```



>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

## What is your favorite candy?

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy %>% select(winpercent)
```

```{r}
candy["Air Heads", ]$winpercent
```
>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

There is a useful `skim()` function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

histogram column

>Q7. What do you think a zero and one represent for the candy$chocolate column?

whether that candy contains chocolate (1) or not (0)
```{r}
candy$chocolate
```

A good place to start any exploratory analysis is with a histogram. You can do this most easily with the base R function `hist()`. Alternatively, you can use `ggplot()` with `geom_hist()`. Either works well in this case and (as always) its your choice.

>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy)+
  aes(x=winpercent)+
  geom_histogram(bins=20)
```

>Q9. Is the distribution of winpercent values symmetrical?

No

```{r}
ggplot(candy)+
  aes(winpercent)+
  geom_density()
```


>Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

fruity

```{r}
#find all choc candy in data set 
choc_inds<- as.logical(candy$chocolate)
choc_candy<- candy[choc_inds,]
# extract winpercent values 
choc.win<-choc_candy$winpercent
#find mean of these values 
choc.mean<-mean(choc.win)

#do the same for fruity candy 
fruity_inds<- as.logical(candy$fruity)
fruit_candy<-candy[fruity_inds,]
fruity.win<-fruit_candy$winpercent
fruity.mean<-mean(fruity.win)

choc.mean
fruity.mean



```


>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win,fruity.win)
```


## Overall Candy Rankings 

>Q13. What are the five least liked candy types in this set?

```{r}
candy %>%
  arrange(desc(winpercent)) %>%
  slice_head(n=5)
```


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>%
  arrange(winpercent) %>%
  slice_head(n=5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.
Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```
Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.


```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
Now, for the first time, using this plot we can answer questions like:
> Q17. What is the worst ranked chocolate candy?

Sixlets

>Q18. What is the best ranked fruity candy?

Starburst 


## Taking a look at Pricepoint 

What about value for money? What is the the best candy for the least money? One way to get at this would be to make a plot of winpercent vs the pricepercent variable. The pricepercent variable records the percentile rank of the candy’s price against all the other candies in the dataset. Lower vales are less expensive and high values more expensive.

To this plot we will add text labels so we can more easily identify a given candy. There is a regular geom_label() that comes with ggplot2. However, as there are quite a few candys in our dataset lots of these labels will be overlapping and hard to read. To help with this we can use the geom_text_repel() function from the ggrepel package.

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reeses Minature

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy),pricepercent)) +
  geom_col()
```

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## Exploring the correlation structure 

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate and fruity

>Q23. Similarly, what two variables are most positively correlated?

chocolate and winpercent


## Principle component analysis 


the main function in base R for this is `prcomp()`and we want to set `scale=TRUE` here

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

lets first look at our main result figure- pc plot PC1 vs PC2 

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```


We can make a much nicer plot with the ggplot2 package but it is important to note that ggplot works best when you supply an input data.frame that includes a separate column for each of the aesthetics you would like displayed in your final plot. To accomplish this we make a new data.frame here that contains our PCA results with all the rest of our candy data. We will then use this for making plots below

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

Again we can use the ggrepel package and the function ggrepel::geom_text_repel() to label up the plot with non overlapping candy names like. We will also add a title and subtitle like so:

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

more candy labels you can change the max.overlaps value to allow more overlapping labels or pass the ggplot object p to plotly like so to generate an interactive plot that you can mouse over to see labels:

```{r}
library(plotly)

#ggplotly(p)
```

Let’s finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of chocolate and fruity and the similar effects of chocolate and bar (i.e. we already know they are correlated).

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are picked up in the positive direction, which makes sense since most hard candy are fruity flavored and come in a box or bag of multiple
