---
title: "Class 16: Obtaining and processing SRA datasets on AWS"
author: "Montserrat (A16536527)"
format: pdf
toc: TRUE
---


##Downstream Analysis 


R function called `tximport()` in the tximport package, which enables straightforward import of Kallisto results


```{r}
# setup the folder and filenames to read
library(tximport)
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.tsv" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```


```{r}
colSums(txi.kallisto$counts)
```

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

Before subsequent analysis, we might want to filter out those annotated transcripts with no reads:

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

And those with no change over the samples:

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

##PCA 


```{r}
pca <- prcomp(t(x), scale=TRUE)
```

```{r}
summary(pca)
```

Now we can use the first two principal components as a co-ordinate system for visualizing the summarized transcriptomic profiles of each sample:

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

PC1 vs PC2

```{r}
library(ggplot2)
condition <- c("control", "control", "treated", "treated")
ggplot(pca$x) + aes(PC1, PC2, col=condition, label=samples)+
  geom_point()+
  geom_text(vjust=-0.5, size=2)
```
PC1 vs PC3 

```{r}
ggplot(pca$x) + aes(PC1, PC3, col=condition, label=samples)+
  geom_point()+
  geom_text(vjust=-0.5, size=2)
```
PC2 vs PC3 

```{r}
ggplot(pca$x) + aes(PC2, PC3, col=condition, label=samples)+
  geom_point()+
  geom_text(vjust=-0.5, size=2)
```

